### 2020-06-16 manuscript/internship_report.Rmd
add conclusion

## Eye Movement Metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007, and @Zemblys2018]. This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics. Belonging to the most basic metrics, which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [similar to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013]. They are specifically distinguished by their change in direction clustered around 180° [@Pekkanen2017]. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].

@@

  p <- ggplot(data, aes(x = true, y = est)) + 
    facet_wrap(vars(par), scales = "free", labeller = label_bquote(.(parnames[[par]]))) +
    geom_point() + geom_smooth(method = "lm") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous(name = "True transition probability") +
    scale_y_continuous(name = "Estimated transition probability")

@@

(ref:plot-regresp-1-2) Regression lines between true and estimated transition response parameters of the two-state HMM in part one. Top facet labels indicate response parameters.

@@

(ref:plot-regresp-1-3) Regression lines between true and estimated transition response parameters of the three-state HMM in part one. Top facet labels indicate response parameters.

@@

(ref:plot-regresp-1-4) Regression lines between true and estimated transition response parameters of the four-state HMM in part one. Top facet labels indicate response parameters.

@@

Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle. The overlap can cause the HMM to confuse parameters and hidden states. The decrease in parameter recovery (especially for shape parameters) due to noise shows that the overlap is enhanced by more dispersion in the data. The shape parameters might be particularly vulnerable to extreme data points. The decrease in state recovery for long and many intervals of missing data can be explained by the tendency of HMMs to stay in the same state when no information is available from the data. However, the decrease was rather small, leading to the conclusion that the HMM can deal well with missing data. Despite these drawbacks, the recovery of the generative model in gazeHMM seems very promising. The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).  

@@

## Advantages of gazeHMM
In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they have only minor influence on the classification and are merely included to compensate for some drawbacks of the generative model. Their default values should be appriate for most applications. A major advantage of gazeHMM is that it does not require human labeled data as input. Instead, it estimates all parameters and hidden states from the data. Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019]. This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data. Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit. Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2017], and even less categorize smooth pursuits [e.g., @Pekkanen2017]. However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events. Therefore, I recommend using gazeHMM with four events onyl for exploratory purposes. Without smooth pursuits, I consider gazeHMM's classification as appropriate for application. Lastly, its implementation in R using depmixS4 [@Visser2010] should make gazeHMM a tool that is easy to use and customize for individual needs.

## Future Directions
Despite its advantages, there several aspects in which gazeHMM can be improved:

Multivariate gamma - conditional independence
Blinks and noise
Non-parametric response distributions
Time varying covariates
Semi-HMM
Metric separating fixations and sp
Constraining parameters

>>>>

## Eye Movement Metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018]. This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics. Belonging to the most basic metrics, which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [similar to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013]. They are specifically distinguished by their change in direction clustered around 180° [@Pekkanen2017]. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].

@@

  p <- ggplot(data, aes(x = true, y = est)) + 
    facet_wrap(vars(par), scales = "free", labeller = label_bquote(.(parnames[[par]]))) +
    geom_point() + geom_smooth(method = "lm") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous(name = "True response parameter") +
    scale_y_continuous(name = "Estimated response parameter")

@@

(ref:plot-regresp-1-2) Regression lines between true and estimated response parameters of the two-state HMM in part one. Top facet labels indicate response parameters.

@@

(ref:plot-regresp-1-3) Regression lines between true and estimated response parameters of the three-state HMM in part one. Top facet labels indicate response parameters.

@@

(ref:plot-regresp-1-4) Regression lines between true and estimated response parameters of the four-state HMM in part one. Top facet labels indicate response parameters.

@@

Considering the results of the simulation study, it seems reasonable that adding the smooth pursuit state to the HMM decreased parameter and state recovery: It is the event that is overlapping most closely with another event (fixations) in terms of velocity, acceleration, and sample-to-sample angle. The overlap can cause the HMM to confuse parameters and hidden states. The decrease in parameter recovery (especially for scale parameters) due to noise shows that the overlap is enhanced by more dispersion in the data. The scale parameters might be particularly vulnerable to extreme data points. The decrease in state recovery for long and many intervals of missing data can be explained by the tendency of HMMs to stay in the same state when no information is available from the data. However, the decrease was rather small, leading to the conclusion that the HMM can deal well with missing data. Despite these drawbacks, the recovery of the generative model in gazeHMM seems very promising. The simulation study gives also an approximate reference for the maximum recovery of hidden states that can be achieved by the HMM (Cohen's kappa values of ~1 for two, ~0.95 for three, and ~0.8 for four events).  

@@

## Advantages of gazeHMM
In view of the four proposed goals that gazeHMM should fulfill, I can draw the following conclusions: Even though gazeHMM does require some parameter settings (in the pre- and postprocessing), they have only minor influence on the classification and are merely included to compensate for some drawbacks of the generative model. Their default values should be appriate for most applications. A major advantage of gazeHMM is that it does not require human labeled data as input. Instead, it estimates all parameters and hidden states from the data. Since human coding is quite laborious, difficult to reproduce, and by times inconsistent [as noted earlier, @Hooge2018], this property makes gazeHMM a good alternative to other recently developed algorithms that require human coded input [@Pekkanen2017; @Zemblys2018; @Bellet2019]. This could also explain why the agreement to human coding is lower for gazeHMM than for algorithms that learn from human labeled data. Another advantage of gazeHMM is its ability to classify four eye movement events, namely fixations, saccades, PSOs, and smooth pursuit. Whereas most algorithms only parse fixations and saccades [@Andersson2017], few are able to classify PSOs [e.g., @Zemblys2017], and even less categorize smooth pursuits [e.g., @Pekkanen2017]. However, including smooth pursuits in gazeHMM led to some undesirable classifications on benchmark data, resulting in rapid switching between fixation and smooth pursuit events. Therefore, I recommend using gazeHMM with four events only for exploratory purposes. Without smooth pursuits, I consider gazeHMM's classification as appropriate for application. Lastly, its implementation in R using depmixS4 [@Visser2010] should make gazeHMM a tool that is easy to use and customize for individual needs.

## Future Directions
Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for an example, see @Balakrishnan2009]. Potential problem of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters). Another option to model the correlation could be to include one of the response variables as a covariate of the other.  
Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred. This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.  
Third, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data. As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements. As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades.  
Fourth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonable short saccades, gazeHMM could explicitly model the duration of events. This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2015]. Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs. This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades. A drawback of the extension would be higher computational costs.  
Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM. This could potentially be achieved by using different response distributions or parameter optimization methods. Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.

## Conclusions
Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events. For smooth pursuits, the classification is not optimal and thus not recommended, yet. On one hand, the algorithm has some advantages over concurrent event classfication algorithms, not relying on human-labeled input being the most important one. On the other hand, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events. These difficulties highlight general challenges for HMMs in general, but also specifically for eye movements. Revisting my research question, I conclude that HMMs are suitable models for classifying eye movement events but still lack the ability to accurately describe them.

*Acknowledgment*. I would like to thank my supervisors Ingmar Visser and Simon Kucharsky for their helpful advice on the development and implementation of the study.
